---
title: "yarraclean: Yarra River Water Quality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{yarraclean}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 4.6
)
```

## Installation

You can install the package directly from GitHub using:

```{r eval = FALSE}
# Install and load the package
remotes::install_github(
  "ETC5523-2025/assignment-4-packages-and-shiny-apps-HendricksonJY",
  build_vignettes = TRUE
)
library(yarraclean)
```
note: ensure you have the remotes package installed.

## Helper packages used inside this vignette

```{r}
#| message: false

library(dplyr)
library(ggplot2)
library(lubridate)
library(tibble)
library(knitr)
```
## Overview

this vignette introduces the ``yarra_river_data`` dataset included in the ``yarraclean`` package.

The dataset has been cleaned, filtered, and reshaped to support: 

- exploratory visualization and analysis
- statistical & predictive modelling
- interactive applications

Each row in the dataset represents a measurement at a monitoring site and time, with variables for pH, turbidity, salinity, total nitrogen, and water termperature. 

```{r}
data("yarra_river_data", package = "yarraclean")
dplyr::glimpse(yarra_river_data)
```

## Data Dictionary

```{r}
#| echo: false

var_dict <- tribble(
~Variable,           ~Type,      ~Units,     ~Description,
"site_id",           "character","—",        "Monitoring site identifier.",
"datetime",          "POSIXct",  "—",        "Timestamp of the observation (local time).",
"ph",                "numeric",  "unitless", "Acidity/alkalinity (0–14). Healthy rivers ~6.5–8.5.",
"salinity_ec25",     "numeric",  "µS/cm",    "Electrical conductivity at 25°C; proxy for salinity/dissolved ions.",
"turbidity",         "numeric",  "NTU",      "Cloudiness from suspended particles; higher = murkier.",
"nitrogen_total",    "numeric",  "mg/L",     "Total nitrogen; nutrient linked to algal growth and eutrophication.",
"water_temperature", "numeric",  "°C",       "Water temperature; affects dissolved oxygen and species tolerance."
)

kable(
var_dict,
align = c("l", "l", "l", "l"),
caption = "Dictionary of variables in `yarra_river_data`."
)

```

## Example: pH by Year (Boxplot)

The following example visualizes pH levels by year using a boxplot. Reference lines at 6.5 and 8.0 indicate the commonly accepted healthy range. 

```{r}
#| echo: false
yarra_river_data |>
dplyr::filter(!is.na(ph), ph != 0) |>
dplyr::mutate(Year = factor(lubridate::year(datetime))) |>
ggplot2::ggplot(ggplot2::aes(x = Year, y = ph)) +
ggplot2::geom_boxplot() +
ggplot2::geom_hline(yintercept = 6.5, linetype = "dashed", color = "red") +
ggplot2::geom_hline(yintercept = 8.0, linetype = "dashed", color = "red") +
ggplot2::geom_vline(xintercept = 4.5, color = "blue") +
ggplot2::labs(
title = "Distribution of pH by Year",
x = "Year", y = "pH"
) +
ggplot2::theme_minimal()

```

## Raw data source

The cleaned dataset in the package is derived from the original raw monitoring data published by the [ Victorian Department of Energy, Environment, and Climate Action](https://data.water.vic.gov.au/WMIS/) via the Water Monitoring Information System (WMIS).

## License

This package is released under the MIT License, which permits reuse, modification, and redistribution with attribution.

The original dataset remains the property of the Victorian Government and is redistributed here only in cleaned form for educational purposes.
